---
title: "Visualizing data distributions"
output: html_notebook
---

# 9 Visualizing data distributions
##
Our first data visualization building block is learning to summarize lists of factors or numeric vectors.
Once a vector has been summarized as a distribution, there are several data visualization techniques to effectively relay this information.
In this Chapter, we first discuss properties of a variety of distributions and how to visualize distributions

## 9.1 Variable types.

We will be working with two types of variables: categorical and numeric. Each can be divided into two other groups: categorical can be ordinal or not, whereas numerical variables can be discrete or continuous.
Keep in mind that discrete numeric data can be considered ordinal. 

## 9.2 Case study: describing student heights

We will use data of hights of students.
```{r}
library(tidyverse)
library(dslabs)
data(heights)
head(heights)
```

## 9.3 Distribution function

The most basic statistical summary of a list of objects or numbers is its distribution.
When there are more categories, then a simple barplot describes the distribution.
When the data is numerical, the task of displaying distributions is more challenging.

## 9.4 Cumulative distribution functions

Numerical data that are not categorical also have distributions. The more useful way to define a distribution for numeric data is to define a function that reports the proportion of the data below a for all possible values of a. This function is called the cumulative distribution function (CDF).
F(a) = P(x <= a)
'Empirical' is added to make the distinction when data is used.

## 9.5 Histograms

The simplest way to make a histogram is to divide the span of our data into non-overlapping bins of the same size. Then, for each bin, we count the number of values that fall in that interval.
What information do we lose? Note that all values in each interval are treated the same when computing bin heights

## 9.6 Smoothed density

The smooth density is basically the curve that goes through the top of the histogram bars when the bins are very, very small.
We can actually control the smoothness of the curve that defines the smooth density through an option in the function that computes the smooth density curve.
While the histogram is an assumption-free summary, the smoothed density is based on some assumptions.

### 9.6.1 Interpreting the y-axis

Note that interpreting the y-axis of a smooth density plot is not straightforward. The best way to determine the proportion of data in that interval is by computing the proportion of the total area contained in that interval.

### 9.6.2 Densities permit stratification

We point out that an advantage of smooth densities over histograms for visualization purposes is that densities makes it easier to compare two distributions.

## 9.7 Exercises

1. In the `murders` dataset, the region is a categorical variable and the following is its distribution:
https://rafalab.github.io/dsbook/book_files/figure-html/barplot-exercise-1.png

To the closet 5%, what proportion of the states are in the North Central region?
About 24%

2. Which of the following is true: B

A. The graph above is a histogram.
-> It's a barplot.

B. The graph above shows only four numbers with a bar plot.
-> Correct

C. Categories are not numbers, so it does not make sense to graph the distribution.
-> We can use barplot to visualize categorical data.

D. The colors, not the height of the bars, describe the distribution.
-> The colors do not have any meanings.

3. The plot below shows the eCDF for male heights:
https://rafalab.github.io/dsbook/book_files/figure-html/ecdf-exercise-1.png

Based on the plot, what percentage of males are shorter than 75 inches?

B. 95%

4. To the closest inch, what height `m` has the property that 1/2 of the male students are taller than `m` and 1/2 are shorter?

C. 69 inches

5. Here is an eCDF of the murder rates across states:
https://rafalab.github.io/dsbook/book_files/figure-html/ecdf-exercise-2-1.png

Knowing that there are 51 states (counting DC) and based on this plot, how many states have murder rates larger than 10 per 100,000 people?

A. 1

6. Based on the eCDF above, which of the following statements are true: D

A. About half the states have murder rates above 7 per 100,000 and the other half below.
-> The murder rate is lower than 7 for most of states(96%).

B. Most states have murder rates below 2 per 100,000.
-> Only about 25% of the states have murder rates below 2.

C. All the states have murder rates above 2 per 100,000.
-> There are some states have murder rates below 2.

D. With the exception of 4 states, the murder rates are below 5 per 100,000.
-> Correct. There is 4 escalations in the range of a>5.

7. Below is a histogram of male heights in our `heights` dataset:
https://rafalab.github.io/dsbook/book_files/figure-html/height-histogram-exercise-1.png

Based on this plot, how many males are between 63.5 and 65.5?

C. 34

8. About what percentage are shorter than 60 inches?

A. 1%

9. Based on the density plot below, about what proportion of US states have populations larger than 10 million?
https://rafalab.github.io/dsbook/book_files/figure-html/density-exercise-1.png

B. 0.15

10. Below are three density plots. Is it possible that they are from the same dataset?
https://rafalab.github.io/dsbook/book_files/figure-html/density-exercise-2-1.png

Which of the following statements is true: D

A. It is impossible that they are from the same dataset.
-> We can plot differently as we change the degree of smoothness.

B. They are from the same dataset, but the plots are different due to code errors.
-> We can plot them on purpose.

C. They are the same dataset, but the first and second plot undersmooth and the third oversmooths.
-> The first one might be oversmooth.

D. They are the same dataset, but the first is not in the log scale, the second undersmooths and the third oversmooths.
-> Correct.

## 9.8 The normal distribution

We often see the average and standard deviation used as summary statistics.
The normal distribution, also known as the bell curve and as the Gaussian distribution, is one of the most famous mathematical concepts in history.
Rather than using data, the normal distribution is defined with a mathematical formula. For any interval (a, b), the proportion of values in that interval can be computed using a formula.
The distribution is symmetric, centered at the average, and most values (about 95%) are within 2 SDs from the average.
If we store the height of males to x:
```{r}
index <- heights$sex=="Male"
x <- heights$height[index]
```

The average of x is:
```{r}
m <- sum(x) / length(x)
```

And the standard deviation of x is:
```{r}
s <- sqrt(sum((x-m)^2) / length(x))
```

Also, we can use pre-bulit functions:
```{r}
m <- mean(x)
s <- sd(x)
c(average = m, sd = s)
```

The normal distribution does appear to be quite a good approximation here.

## 9.9 Standard units

We now will see how well this approximation works at predicting the proportion of values within intervals. For data that is approximately normally distributed, it is convenient to think in terms of standard units as `z = (x - m)/s`.
In R, we can obtain standard units using the function `scale`:
```{r}
z <- scale(x)
```

Now to see how many men are within 2 SDs from the average, we simply type:
```{r}
mean(abs(z) < 2)
```

## 9.10 Quantile-quantile plots

A systematic way to assess how well the normal distribution fits the data is to check if the observed and predicted proportions match. In general, this is the approach of the quantile-quantile plot (QQ-plot).
Φ(x)(we call it phi) is the area(=provability) of the interval from negative infinity to x. In R, we can evaluate Φ using the `pnorm` function:
```{r}
pnorm(-1.96)
```

The inversion of Φ(Φ^-1^(x)) gives us the theoretical quantiles for the normal distribution. In R, we can evaluate the inverse of Φ using the `qnorm` function:
```{r}
qnorm(0.975)
```

In these cases, the x was the standard normal distribution by default (mean = 0, standard deviation = 1). But we can use any normal distributions by using arguments 'mean' and 'sd':
```{r}
qnorm(0.975, mean = 5, sd = 2)
```

For the normal distribution, all the calculations related to quantiles are done without data, thus the name theoretical quantiles.
Using R code, we can define `q` as the value for which `mean(x <= q) = p`:
```{r}
mean(x <= 69.5)
```

To construct a QQ-plot, we do the following:
1. Define a vector of m proportions p~1~,p~2~,…,p~m~.
2. Define a vector of quantiles q~1~,…,q~m~ for your data for the proportions p~1~,…,p~m~. We refer to these as the sample quantiles.
3. Define a vector of theoretical quantiles for the proportions p~1~,…,p~m~ for a normal distribution with the same average and standard deviation as the data.
4. Plot the sample quantiles versus the theoretical quantiles.

For R code, first we define the vector of propotions:
```{r}
p <- seq(0.05, 0.95, 0.05)
```

To obtain the quantiles from the data, we can use the function `quantile` like this:
```{r}
sample_quantiles <- quantile(x, p)
```

To obtain the theoretical normal distribution quantiles with the corresponding average and SD, we use the` function `qnorm:
```{r}
theoretical_quantiles <- qnorm(p, mean = mean(x), sd = sd(x))
```

To see if they match or not, we plot them against each other and draw the identity line:
```{r}
qplot(theoretical_quantiles, sample_quantiles) + geom_abline()
```

Notice that this code becomes much cleaner if we use standard units:
```{r}
# sample_quantiles <- quantile(z, p)
# theoretical_quantiles <- qnorm(p) 
# qplot(theoretical_quantiles, sample_quantiles) + geom_abline()
```

And if we use the function `geom_qq` from ggplot2:
```{r}
heights %>% filter(sex=="Male") %>%
  ggplot(aes(sample = scale(height))) + 
  geom_qq() +
  geom_abline()
```

## 9.11 Percentiles

Percentiles are special cases of quantiles that are commonly used. The percentiles are the quantiles you obtain when setting the p at 0.01,0.02,...,0.99 and we call the case of p = n, the n^th^ percentile.
If we set p = 0.25, 0.50, 0.75, we call that as quartiles.

## 9.12 Boxplots

Suppose we want to summarize the murder rate distribution, we can quickly see that the normal approximation does not apply here:
https://rafalab.github.io/dsbook/book_files/figure-html/hist-qqplot-non-normal-data-1.png

Tukey further suggested that we ignore outliers when computing the range and instead plot these as independent points. He suggested we plot these numbers as a “box”
The box defined by the 25% and 75% percentile and the whiskers showing the range. The distance between these two is called the interquartile range. The two points are outliers according to Tukey’s definition. The median is shown with a horizontal line. Today, we call these boxplots.

## 9.13 Stratification(계층화)

In data analysis we often divide observations into groups based on the values of one or more variables associated with those observations.
Stratification is common in data visualization because we are often interested in how the distribution of variables differ across different subgroups.

## 9.14 Case study: describing student heights (continued)

Using the histogram, density plots and QQ-plots, we have become convinced that the male height data is well approximated with a normal distribution.
We learned that boxplots are useful when we want to quickly compare two or more distributions. Here are the heights for men and women.
We expect that they will follow a normal distribution, just like males. However, exploratory plots reveal that the approximation is not as useful.
Finally, we also see five points in the QQ-plot that suggest shorter than expected heights for a normal distribution.
In any case, data visualization has helped discover a potential flaw in our data.
Regarding the five smallest values, note that these values are:
```{r}
heights %>%
  filter(sex=="Female") %>%
  top_n(5, desc(height)) %>%
  pull(height)
```

## 9.15 Exercises

1. Define variables containing the heights of males and females like this:
```{r}
library(dslabs)
data(heights)
male <- heights$height[heights$sex=="Male"]
female <- heights$height[heights$sex=="Female"]
```
How many measurements do we have for each?

```{r}
length(male)
length(female)
```
812 for male and 238 for female.

2. Suppose we can’t make a plot and want to compare the distributions side by side. We can’t just list all the numbers. Instead, we will look at the percentiles. Create a five row table showing `female_percentiles` and `male_percentiles` with the 10th, 30th, 50th, …, 90th percentiles for each sex. Then create a data frame with these two as columns.
```{r}
p <- seq(0.1, 0.9, 0.2)

percentile <- data.frame(male = quantile(male, p),
                         female = quantile(female, p))

percentile
```

3. Study the following boxplots showing population sizes by country:
https://rafalab.github.io/dsbook/book_files/figure-html/boxplot-exercise-1.png

Which continent has the country with the biggest population size?
It's asia.

4. What continent has the largest median population size?
It's africa.

5. What is median population size for Africa to the nearest million?
It's about 10 million.

6. What proportion of countries in Europe have populations below 14 million?

A 0.99

7. If we use a log transformation, which continent shown above has the largest interquartile range?
It`s asia.

8. Load the height data set and create a vector `x` with just the male heights:
```{r}
library(dslabs)
data(heights)
x <- heights$height[heights$sex=="Male"]
```
What proportion of the data is between 69 and 72 inches (taller than 69, but shorter or equal to 72)? Hint: use a logical operator and `mean`.
```{r}
mean(x > 69 & x <= 72)
```

9. Suppose all you know about the data is the average and the standard deviation. Use the normal approximation to estimate the proportion you just calculated. Hint: start by computing the average and standard deviation. Then use the `pnorm` function to predict the proportions.
```{r}
m = mean(x)
sd = sd(x)
pnorm(72, m, sd) - pnorm(69, m, sd)
```

10. Notice that the approximation calculated in question two is very close to the exact calculation in the first question. Now perform the same task for more extreme values. Compare the exact calculation and the normal approximation for the interval (79,81]. How many times bigger is the actual proportion than the approximation?
```{r}
a <- mean(x > 79 & x <= 81)
b <- pnorm(81, m, sd) - pnorm(79, m, sd)
a
b
a/b
```

11. Approximate the distribution of adult men in the world as normally distributed with an average of 69 inches and a standard deviation of 3 inches. Using this approximation, estimate the proportion of adult men that are 7 feet tall or taller, referred to as seven footers. Hint: use the `pnorm` function.
```{r}
1 - pnorm(84, m, sd)
```

12. There are about 1 billion men between the ages of 18 and 40 in the world. Use your answer to the previous question to estimate how many of these men (18-40 year olds) are seven feet tall or taller in the world?
```{r}
a <- round((1 - pnorm(84, m, sd)) * 10^9)
a
```

13. There are about 10 National Basketball Association (NBA) players that are 7 feet tall or higher. Using the answer to the previous two questions, what proportion of the world’s 18 to 40 year old seven footers are in the NBA?
```{r}
10/a
```

14. Repeat the calculations performed in the previous question for Lebron James’ height: 6 feet 8 inches. There are about 150 players that are at least that tall.
```{r}
1- pnorm(78, m, sd)
150 / round((1- pnorm(78, m, sd)) * 10^9)
```

15. In answering the previous questions, we found that it is not at all rare for a seven footer to become an NBA player. What would be a fair critique of our calculations:

A. Practice and talent are what make a great basketball player, not height.
-> I do not know.

B. The normal approximation is not appropriate for heights.
-> Maybe or not. But for this sample, we can use that approximation.

C. As seen in question 3, the normal approximation tends to underestimate the extreme values. It’s possible that there are more seven footers than we predicted.
-> Right.

D. As seen in question 3, the normal approximation tends to overestimate the extreme values. It’s possible that there are less seven footers than we predicted.
-> Reverse.

## 9.16 ggplot2 geometries

### 9.16.1 Barplots

To generate a barplot we can use the `geom_bar` geometry:
```{r}
murders %>%
  ggplot(aes(region)) +
  geom_bar()
```

We often already have a table with a distribution that we want to present as barplot:
```{r}
data(murders)
tab <- murders %>% 
  count(region) %>% 
  mutate(proportion = n/sum(n))
tab
```

And we can change with it:
```{r}
tab %>%
  ggplot(aes(region, proportion)) +
  geom_bar(stat = "identity")
```

### 9.16.2 Histograms

To generate histograms we use `geom_histogram`.
```{r}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) + 
  geom_histogram()
```

Then we get a message
> `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

So:
```{r}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) + 
  geom_histogram(binwidth = 1)
```
And we give some more labels like title and color and so on:
```{r}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) +
  geom_histogram(binwidth = 1, fill = "blue", col = "black") +
  xlab("Male heights in inches") + 
  ggtitle("Histogram")
```

### 9.16.3 Density plots

To create a smooth density, we use the `geom_density`:
```{r}
heights %>% 
  filter(sex == "Female") %>%
  ggplot(aes(height)) +
  geom_density()
```

To fill in with color, we can use the argument `fill`:
```{r}
heights %>% 
  filter(sex == "Female") %>%
  ggplot(aes(height)) +
  geom_density(fill="blue")
```
To change the smoothness of the density, we use the argument `adjust`:
```{r}
heights %>% 
  filter(sex == "Female") %>%
  ggplot(aes(height)) +
  geom_density(fill = "blue", adjust = 2)
```

### 9.16.4 Boxplots

The geometry for boxplot is `geom_boxplot`:
```{r}
heights %>%
  ggplot(aes(sex, height)) +
  geom_boxplot()
```

### 9.16.5 QQ-plots

For qq-plots we use the geometry `geom_qq`:
```{r}
heights %>%
  filter(sex=="Male") %>%
  ggplot(aes(sample = height)) +
  geom_qq()
```

By default, the sample variable is compared to a normal distribution with average 0 and standard deviation 1. To change this, we use the argument `dparams`. Adding an identity line is as simple as assigning another layer. For straight lines, we use the function `geom_abline`.
```{r}
params <- heights %>%
  filter(sex=="Male") %>%
  summarize(mean = mean(height), sd = sd(height))

heights %>%
  filter(sex=="Male") %>%
  ggplot(aes(sample = height)) +
  geom_qq(dparams = params) +
  geom_abline()
```

Another option here is to scale the data first and then make a qqplot against the standard normal.
```{r}
heights %>% 
  filter(sex=="Male") %>%
  ggplot(aes(sample = scale(height))) + 
  geom_qq() +
  geom_abline()
```

### 9.16.6 Images

Images were not needed for the concepts described in this chapter, but we will use images at later. The two geometries used to create images: `geom_tile` and `geom_raster`.

For example, here is a data frame about `x` and `y`:
```{r}
x <- expand.grid(x = 1:12, y = 1:10) %>% 
  mutate(z = 1:120) 
```
It is a tiyd version of `matrix(1:120, 12, 10)`

To plot the image we use the following code:
```{r}
x %>% ggplot(aes(x, y, fill = z)) + 
  geom_raster()
```

With these images you will often want to change the color scale. This can be done through the `scale_fill_gradientn layer`.
```{r}
x %>%
  ggplot(aes(x, y, fill = z)) + 
  geom_raster() + 
  scale_fill_gradientn(colors =  terrain.colors(10))
```

### 9.16.7 Quick plots

Suppose we have the female heights in an object `x`:
```{r}
x <- heights %>% 
  filter(sex=="Male") %>% 
  pull(height)
```
To make a quick histogram we can use:
```{r}
qplot(x)
```

To make a quick qqplot you have to use the argument `sample`:
```{r}
qplot(sample = scale(x)) +
  geom_abline()
```

If we supply a factor and a numeric vector, we obtain a plot like the below. Note that in the code below we are using the argument `data`:
```{r}
heights %>% qplot(sex, height, data = .)
```
We can also select a specific geometry by using the argument `geom`:
```{r}
heights %>% qplot(sex, height, data = ., geom = "boxplot")
```

We can also use the argument `geom` to generate a density plot instead of a histogram:
```{r}
qplot(x, geom = "density")
```

Although not as much as with `ggplot`, we do have some flexibility to improve the results of `qplot`:
```{r}
qplot(x, bins=15, color = I("black"), xlab = "Population")
```

## 9.17 Exercises

1. Now we are going to use the `geom_histogram` function to make a histogram of the heights in the `height` data frame. When reading the documentation for this function we see that it requires just one mapping, the values to be used for the histogram. Make a histogram of all the plots.

What is the variable containing the heights?

D. heights$height

2. Now create a ggplot object using the pipe to assign the heights data to a ggplot object. Assign `height` to the x values through the `aes` function.
```{r}
heights %>% ggplot(aes(height))
```

3. Now we are ready to add a layer to actually make the histogram. Use the object created in the previous exercise and the `geom_histogram` function to make the histogram.
```{r}
heights %>% ggplot(aes(height)) +
  geom_histogram()
```

4. Note that when we run the code in the previous exercise we get the warning: `stat_bin() using bins = 30`. Pick better value with `binwidth`.

Use the `binwidth` argument to change the histogram made in the previous exercise to use bins of size 1 inch.
```{r}
heights %>% ggplot(aes(height)) +
  geom_histogram(binwidth = 1)
```

5. Instead of a histogram, we are going to make a smooth density plot. In this case we will not make an object, but instead render the plot with one line of code. Change the geometry in the code previously used to make a smooth density instead of a histogram.
```{r}
heights %>% ggplot(aes(height)) +
  geom_density()
```

6. Now we are going to make a density plot for males and females separately. We can do this using the `group` argument. We assign groups via the aesthetic mapping as each point needs to a group before making the calculations needed to estimate a density.
```{r}
heights %>% ggplot(aes(height, group = sex)) +
  geom_density()
```

7. We can also assign groups through the `color` argument. This has the added benefit that it uses color to distinguish the groups. Change the code above to use color
```{r}
heights %>% ggplot(aes(height, group = sex, color = sex)) +
  geom_density()
```

8. We can also assign groups through the `fill` argument. This has the added benefit that it uses colors to distinguish the groups, like this:
```{r}
heights %>% 
  ggplot(aes(height, fill = sex)) + 
  geom_density() 
```
However, here the second density is drawn over the other. We can make the curves more visible by using alpha blending to add transparency. Set the `alpha` parameter to 0.2 in the `geom_density` function to make this change.
```{r}
heights %>% 
  ggplot(aes(height, fill = sex)) + 
  geom_density(alpha = 0.2) 
```