---
title: "Assignment 0918"
output: html_notebook
---
# 5. The tidyverse
## 5.0
We can load tidyverse packages!
```{r}
library(tidyverse)
```

And we can use dplyr, purrr, ggplot, readr and something else from now on.

## 5.1 Tidy data

We can find 'tidy' dataset like 'murders':
```{r}
library(dslabs)
data("murders")
head(murders)
```

But we can find that some datasets are not that tidy. Like data of fertility that ordered only by year not by country.
If we tide that data with order of year and country both, than we can understand that better.

## 5.2 Exercises

1. Examine the built-in dataset 'co2'. Which of the following is true:
```{r}
data(co2)
head(co2)
```
A. 'co2' is tidy data: it has one year for each row.
-> I cannot find the time that the data are measured.

B. 'co2' is not tidy: we need at least one column with a character vector. 
-> We do not be forced to use character vector to tidy datasets when we can find the purpose of the vector.

C. 'co2' is not tidy: it is a matrix instead of a data frame.
-> That is a matrix composed of one vector.

D. 'co2' is not tidy: to be tidy we would have to wrangle it to have three columns (year, month and value), then each co2 observation would have a row.
-> Correct.

2. Examine the built-in dataset 'ChickWeight'. Which of the following is true:
```{r}
data("ChickWeight")
ChickWeight
```

A. 'ChickWeight' is not tidy: each chick has more than one row.
-> Correct

B. 'ChickWeight' is tidy: each observation (a weight) is represented by one row. The chick from which this measurement came from is one the variables.
-> The description above is the reason for that the dataset is not tidy.

C. 'ChickWeight' is not a tidy: we are missing the year column.
-> For this, it seems needless to insert the year column.

D. 'ChickWeight' is tidy: it is stored in a data frame.
-> Not all data frames are tidy. It depends on the structure of them.

3. Examine the built-in dataset 'BOD'. Which of the following is true:
```{r}
data("BOD")
BOD
```

A. 'BOD' is not tidy: it only has six rows.
-> We cannot decide the dataset is tidy with the number of rows in that.

B. 'BOD' is not tidy: the first column is just an index.
-> It looks really tidy from my point of view.

C. 'BOD' is tidy: each row is an observation with two values (time and demand)
-> Correct

D. 'BOD' is tidy: all small datasets are tidy by definition.
-> Same as A

4. Which of the following built-in datasets is tidy (you can pick more than one):

A. BJsales
```{r}
data(BJsales)
head(BJsales)
BJsales
```
-> Okay. It seems tidy.

B. EuStockMarkets
```{r}
data(EuStockMarkets)
head(EuStockMarkets)
EuStockMarkets
```
-> Good. It seems tidy.

C. DNase
```{r}
data(DNase)
head(DNase)
```
-> It is not tidy. I want that data would be ordered by concentration.

D. Formaldehyde
```{r}
data(Formaldehyde)
head(Formaldehyde)
```
-> Good. Each rows have two data and we can easily find the relation between them.

E. Orange
```{r}
data(Orange)
head(Orange)
```
-> Not tidy. I want that would be ordered by age.

F. UCBAdmissions
```{r}
data(UCBAdmissions)
head(UCBAdmissions)
UCBAdmissions
```
-> Good, It looks tidy. But it will be better that the category of Dept gets the first priority. (Dept->Admit->Gender)

## 5.3 Manipulating data frames

The package 'dplyr' from tidyverse introduces functions that perform some of the most common operations when working with data frames.

### 5.3.1 Adding a column with 'mutate'

If we want to add another column like result of caculation between other two column, we use 'mutate':
```{r}
library(dslabs)
data("murders")
murders <- mutate(murders, rate = total / population * 100000)
```

It does not crash even though we use undefined arguments like 'total' or 'population' because we called a data frame 'murders' and the function in dplyr can process data frame with own variables.

And we can see;
```{r}
head(murders)
```
the new colunm is added.

### 5.3.2 Subsetting with 'filter'

If we want to only show the rows for which the murder rate is lower than 0.71;
```{r}
filter(murders, rate <= 0.71)
```
we can use function 'filter'.

### 5.3.3 Selecting columns with 'select'

And we can only show the columns that we want to show;
```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= 0.71)
```
with using function 'select'

## 5.4 Exercises

1. Load the dplyr package and the murders dataset.
```{r}
library(dplyr)
library(dslabs)
data(murders)
```
You can add columns using the dplyr function mutate. This function is aware of the column names and inside the function you can call them unquoted:
```{r}
murders <- mutate(murders, population_in_millions = population / 10^6)
```
We can write 'population' rather than 'murders$population'. The function 'mutate' knows we are grabbing columns from 'murders'.

Use the function 'mutate' to add a murders column named 'rate' with the per 100,000 murder rate as in the example code above. Make sure you redefine 'murders' as done in the example code above ( murders <- [your code]) so we can keep using this variable.
```{r}
murders <- mutate(murders, rate = total / population * 100000)
head(murders)
```

2. If 'rank(x)' gives you the ranks of 'x' from lowest to highest, 'rank(-x)' gives you the ranks from highest to lowest. Use the function 'mutate' to add a column 'rank' containing the rank, from highest to lowest murder rate. Make sure you redefine 'murders' so we can keep using this variable.
```{r}
murders <- mutate(murders, rank = rank(-murders$rate))
head(murders)
```

3. With dplyr, we can use 'select' to show only certain columns. For example, with this code we would only show the states and population sizes:
```{r}
select(murders, state, population) %>% head()
```
Use 'select' to show the state names and abbreviations in 'murders'. Do not redefine 'murders', just show the results.
```{r}
select(murders, state, abb)
```

4. The dplyr function 'filter' is used to choose specific rows of the data frame to keep. Unlike 'select' which is for columns, 'filter' is for rows. For example, you can show just the New York row like this:
```{r}
filter(murders, state == "New York")
```
You can use other logical vectors to filter rows.

Use 'filter' to show the top 5 states with the highest murder rates. After we add murder rate and rank, do not change the murders dataset, just show the result. Remember that you can filter based on the 'rank' column.
```{r}
filter(murders, rank <= 5)
```

5. We can remove rows using the '!=' operator. For example, to remove Florida, we would do this:
```{r}
no_florida <- filter(murders, state != "Florida")
```
Create a new data frame called 'no_south' that removes states from the South region. How many states are in this category? You can use the function 'nrow' for this.
```{r}
no_south <- filter(murders, region != "South")
nrow(no_south)
```

6. We can also use '%in%' to filter with dplyr. You can therefore see the data from New York and Texas like this:
```{r}
filter(murders, state %in% c("New York", "Texas"))
```
Create a new data frame called 'murders_nw' with only the states from the Northeast and the West. How many states are in this category?
```{r}
murders_nw <- filter(murders, region %in% c('Northeast', 'West'))
murders_nw
nrow(murders_nw)
```

7. Suppose you want to live in the Northeast or West and want the murder rate to be less than 1. We want to see the data for the states satisfying these options. Note that you can use logical operators with 'filter'. Here is an example in which we filter to keep only small states in the Northeast region.
```{r}
filter(murders, population < 5000000 & region == "Northeast")
```
Make sure 'murders' has been defined with 'rate' and 'rank' and still has all states. Create a table called 'my_states' that contains rows for states satisfying both the conditions: it is in the Northeast or West and the murder rate is less than 1. Use 'select' to show only the state name, the rate and the rank.
```{r}
my_states <- select(filter(murders, rate < 1 & region %in% c("Northeast", "West")), state, rate, rank)
my_states
```


## 5.5 The pipe: '%>%'

'%>%' is tbe pipe operator that sends the result data before to another function after:
```{r}
murders %>% select(state, region, rate) %>% filter(rate <= 0.71)
```

And for example:
```{r}
16 %>% sqrt()
```
The data sends to right function.

And;
```{r}
16 %>% sqrt() %>% log2()
```
we can continue several pipes.

The pipe operator send data as the first argument:
```{r}
16 %>% sqrt() %>% log(base = 2)
```

As we use the pipe operators, we can simplify codes:
```{r}
murders %>% select(state, region, rate) %>% filter(rate <= 0.71)
```

## 5.6 Exercises

1. The pipe '%>%' can be used to perform operations sequentially without having to define intermediate objects. Start by redefining murder to include rate and rank.
```{r}
murders <- mutate(murders, rate =  total / population * 100000, rank = rank(-rate))
```
In the solution to the previous exercise, we did the following:
```{r}
my_states <- filter(murders, region %in% c("Northeast", "West") & rate < 1)
select(my_states, state, rate, rank)
```
The pipe '%>%' permits us to perform both operations sequentially without having to define an intermediate variable 'my_states'. We therefore could have mutated and selected in the same line like this:
```{r}
mutate(murders, rate =  total / population * 100000, rank = rank(-rate)) %>% 
  select(state, rate, rank)
```
Notice that 'select' no longer has a data frame as the first argument. The first argument is assumed to be the result of the operation conducted right before the '%>%'.

Repeat the previous exercise, but now instead of creating a new object, show the result and only include the state, rate, and rank columns. Use a pipe '%>%' to do this in just one line.
```{r}
murders %>% select(state, rate, rank)
```

2. Reset 'murders' to the original table by using 'data(murders)'. Use a pipe to create a new data frame called 'my_states' that considers only states in the Northeast or West which have a murder rate lower than 1, and contains only the state, rate and rank columns. The pipe should also have four components separated by three %>%. The code should look something like this:
```{r}
data(murders)
my_states <- murders %>%
  mutate(rate = total / population * 100000, rank = rank(-rate)) %>% 
  filter(rate < 1 & region %in% c('Northeast', 'West')) %>% 
  select(state, rate, rank)
```